# Machine Learning (Course Project)
**Data Science -> Practical Machine Learning -> Peer Assessment**  
<br><br>  

```{r Setup, echo=FALSE, results='hide'}
# Setup
require(knitr)
require(tools)
opts_chunk$set(echo=FALSE, results="markup")
opts_knit$set(verbose=TRUE)
```

-------------------
### Executive Summary  


-------------------
### Data Loading, Preprocessing and Exploratory Analysis   

We'll start by downloading the training and testing cases from the given assignment URLs. Once we make sure we have the right files, they are loaded into memory.

```{r Loading}
# Loading
kRemote <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
kFile   <- "train.csv"
kMd5    <- "56926c78af383dcdc2060407942e52e9"
    
if (!file.exists(kFile)){
    # File is not present at the working directory. Lets download it!
    method <-"auto"  # Default method: shall be fine for MS Windows (untested!)
    if (.Platform$OS.type=="unix") method <- "curl"  # Use for unix-like systems
    download.file(kRemote, kFile, method, FALSE, "wb")
}

if (!file.exists(kFile)){
    stop("Download failed!")
} else {
    # File is now present (it was before or it was sucessfully downloaded).
    # Lets check if it is the expected file and no curruption ocurred during 
    # download. I'll compare it's MD5 hash with a precomputed one.
    if (as.vector(md5sum(kFile))!=kMd5){
        stop("File is not correct!")
    } else {
        # The correct file shall be present. 
        trainData <- read.csv(kFile, stringsAsFactors=FALSE)
        #print("Training Data sucessfully loaded!")
    }
}

kRemote <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
kFile   <- "test.csv"
kMd5    <- "bc4174f3ec5dfcc5c570a1d2709272d9"

if (!file.exists(kFile)){
    method <-"auto"
    if (.Platform$OS.type=="unix") method <- "curl"
    download.file(kRemote, kFile, method, FALSE, "wb")
}

if (!file.exists(kFile)){
    stop("Download failed!")
} else {
    if (as.vector(md5sum(kFile))!=kMd5){
        stop("File is not correct!")
    } else {
        testData <- read.csv(kFile, stringsAsFactors=FALSE)
        #print("Testing Data sucessfully loaded!")
    }
}
rm(kRemote, kFile, kMd5)

cat("Training dataset:", dim(trainData), "\nTesting dataset:", dim(testData))

#Preprocessing

```

There are two files with `r dim(trainData)[2]` variables and `r dim(trainData)[1]` records for the training dataset and `r dim(testData)[1]` records for the test cases. We'll keep the test cases apart until the prediction phase, later in this document. 




-------------------
### References

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. _Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13)_ . Stuttgart, Germany: ACM SIGCHI, 2013.